Fractional Types

### 2016-12-19 typespp.tex
edit conclusion (and insert HoTT reference earlier)

\noindent Both classes of programs, 1-combinators $c$, and
2-combinators~$\alpha$, denote \emph{equivalences} in the HoTT sense. The
elements $c$ of 1-combinators denote type isomorphisms. The elements $\alpha$ of
2-combinators denote the set of sound and complete equivalences between these
type isomorphisms. Using the 1-combinators, it is possible to write any
reversible boolean function and hence encode arbitrary boolean functions by a
technique that goes back to Toffoli~\cite{Toffoli:1980}. The 2-combinators
provide a layer of programs that computes semantics-preserving transformations
of 1-combinators. As a small example, let us abbreviate $\ot \oplus \ot$ as the
type $\mathbb{2}$ of booleans and examine two possible implementations of
boolean negation. The first directly uses the primitive combinator
$\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to exchange the two
values of type~$\mathbb{2}$; the second use three consecutive uses of $\swapp$
to achieve the same effect:
\[\begin{array}{rcl}
\mathsf{not_1} &=& \swapp \\
\mathsf{not_2} &=& (\swapp \odot \swapp) \odot \swapp
\end{array}\]
We can write a 2-combinator whose \emph{type} is $\mathsf{not_2}
\isotwo \mathsf{not_1}$:
\[
(\linvdl ~\respstwo~ \idisotwo)~\transtwo~\idldl
\]
which not only shows the equivalence of the two implementations of negation but
also shows \emph{how} to transform one to the other. This rewriting focuses
on the first two occurrences of $\swapp$ and uses $\linvdl$ to reduce them to
$\idiso$ since they are inverses. It then uses $\idldl$ to simplify the
composition of $\idiso$ with $\swapp$ to just $\swapp$.

@@

We have presented a natural notion of \emph{fractional types} that enriches a
class of reversible programming languages in several dimensions.  Further
research might show how to use fractional types in a conventional (i.e.,
irreversible) programming language, their full potential is only achieved when
the ambient language guarantees that no information is created or erased. The
key semantic insight is that iterating a reversible program $p$ on a finite type
must eventually reach the identity in $\ord{p}$ steps. By being careful not to
collapse proofs, each such reversible program has $\ord{p}$ distinct proofs of
reversibility and hence gives rise to a groupoid with cardinality
$\frac{1}{\ord{p}}$. Going from this observation to a full programming language
required several difficult and subtle design choices which we have explored to
produce $\Pi^/$. The latter language has a natural denotational semantics where
types denote groupoids. Its operational semantics requires a mechanism to
express a computational effect which enables spatially separated parts of the
program to communicate in a way that is reminiscent of entanglement in quantum
mechanics. It appears possible to realize such an operational semantics using
global reference cells, backtracking, or other conventional technique.

Our fractional types extend the natural denotation of types from sets to
non-trivial groupoids but they only scratch the surface of the tower of weak
$\omega$-groupoids that is expressible in HoTT. A long term goal of our research
is to find natural type constructors inspired by the rich combinatorial
structure of weak $\omega$-groupoids and that provide novel programming
abstractions. We conclude by outlining several interesting connections and
potential avenues for future work.

@@

\paragraph*{Conservation of Information and Negative Entropy.} According to the
conventional theory of communication~\cite{Shannon1948}, a type with~$N$ values
is viewed as an abstract system that has~$N$ distinguishable states where the
amount of information contained in each state is $(\log{N})$. This entropy is a
measure of information which materializes itself in memory or bandwidth
requirements when storing or transmitting elements of this type. Thus a type
with 8 elements needs 3 bits of memory for storage or 3 bits of bandwidth for
communication. The logarithmic map implies that information contained in a
composite state is the sum of the information contained in its constituents. For
example, the type $\mathbb{2} \otimes \mathbb{3}$ can be thought of a composite
system consisting of two independent unrelated subsystems. Each state of the
composite system therefore contains $\log (2 * 3) = \log 2 + \log 3$ bits which
is the sum of the information contained in each subsystem. If quantum field
theory is correct (as it so far seems to be) then information, during any
physical process, is neither created nor
destroyed. Landauer~\cite{Landauer:1961,bennett1985fundamental,Landauer},
Bennet~\cite{Bennett:1973:LRC,bennett2003notes,bennett2010notes},
Fredkin~\cite{fredkin1982conservative} and others made compelling arguments that
this physical principle induces a corresponding computational principle of
``conservation of information.'' In the context of finite types, generated from
the empty type $\zt$, the unit type $\ot$, and sums and products $\oplus$ and
$\otimes$, this principle states that the foundational (i.e., physical) notion
of computation is computation via type
isomorphisms~\cite{James:2012:IE:2103656.2103667} or type
equivalences~\cite{Carette2016}, which are both sound and complete with respect
to cardinality-preserving maps. The introduction, in $\pifrac$, of types
(groupoids) with fractional cardinalities introduces types with \emph{negative
  entropy}. For example, a type with cardinality $\frac{1}{8}$ has entropy
$\log{\frac{1}{8}} = -3$. It is natural to interpret this negative entropy just
like we interpret ``negative money,'' as a debt to be repaid by some other part
of the system. In fact, the zig-zag example can be viewed as modeling a credit
card transaction where the money is produced at the output site by generating a
corresponding debt that is reconciled at the input site.

>>>>

\noindent Both classes of programs, 1-combinators $c$, and
2-combinators~$\alpha$, denote \emph{equivalences} in the Homotopy Type
Theory (HoTT) sense~\cite{hottbook}. The
elements $c$ of 1-combinators denote type isomorphisms. The elements $\alpha$ of
2-combinators denote the set of sound and complete equivalences between these
type isomorphisms. Using the 1-combinators, it is possible to write any
reversible boolean function and hence encode arbitrary boolean functions by a
technique that goes back to Toffoli~\cite{Toffoli:1980}. The 2-combinators
provide a layer of programs that computes semantics-preserving transformations
of 1-combinators. As a small example, let us abbreviate $\ot \oplus \ot$ as the
type $\mathbb{2}$ of booleans and examine two possible implementations of
boolean negation. The first directly uses the primitive combinator
$\swapp : \tau_1 \oplus \tau_2 \iso \tau_2 \oplus \tau_1$ to exchange the two
values of type~$\mathbb{2}$; the second use three consecutive uses of $\swapp$
to achieve the same effect:
\[\begin{array}{rcl}
\mathsf{not_1} &=& \swapp \\
\mathsf{not_2} &=& (\swapp \odot \swapp) \odot \swapp
\end{array}\]
We can write a 2-combinator whose \emph{type} is $\mathsf{not_2}
\isotwo \mathsf{not_1}$:
\[
(\linvdl ~\respstwo~ \idisotwo)~\transtwo~\idldl
\]
which not only shows the equivalence of the two implementations of negation but
also shows \emph{how} to transform one to the other. This rewriting focuses
on the first two occurrences of $\swapp$ and uses $\linvdl$ to reduce them to
$\idiso$ since they are inverses. It then uses $\idldl$ to simplify the
composition of $\idiso$ with $\swapp$ to just $\swapp$.

@@

We have presented a natural notion of \emph{fractional types} that enriches a
class of reversible programming languages in several dimensions.  While further
research might show how to use fractional types in a conventional (i.e.,
irreversible) programming language, their full potential is only achieved when
the ambient language guarantees that no information is created or erased. The
key semantic insight is that iterating a reversible program $p$ on a finite type
must eventually reach the identity in $\ord{p}$ steps. By being careful not to
collapse proofs, each such reversible program has $\ord{p}$ distinct proofs of
reversibility and hence gives rise to a groupoid with cardinality
$\frac{1}{\ord{p}}$. Going from this observation to a full programming language
required several difficult and subtle design choices which we have explored to
produce $\Pi^/$. As with most functional languages, it has 
a natural denotational semantics where
types denote groupoids. Its operational semantics requires a mechanism to
express a computational effect which enables spatially separated parts of the
program to communicate in a way that is reminiscent of entanglement in quantum
mechanics. It appears possible to realize such an operational semantics using
global reference cells, backtracking, or other conventional technique.  Rather
than going via effects, we chose dependent types to track the required
constraints via dataflow.  Of our various attempts at an operational semantics,
this one turned out simplest.

Our fractional types have natural denotations which are non-trivial groupoids,
but they only scratch the surface of the tower of weak $\omega$-groupoids that
is expressible in HoTT~\cite{hottbook}. A long term goal of our research
is to find natural type constructors inspired by the rich combinatorial
structure of weak $\omega$-groupoids that provide novel programming
abstractions. We conclude by outlining several interesting connections and
potential avenues for future work.

@@

\paragraph*{Conservation of Information and Negative Entropy.} According to the
conventional theory of communication~\cite{Shannon1948}, a type with~$N$ values
is viewed as an abstract system that has~$N$ distinguishable states where the
amount of information contained in each state is $(\log{N})$. This entropy is a
measure of information which materializes itself in memory or bandwidth
requirements when storing or transmitting elements of this type. Thus a type
with $8$ elements needs $3$ bits of memory for storage or $3$ bits of bandwidth
for communication. The logarithmic map implies that information contained in a
composite state is the sum of the information contained in its constituents. For
example, the type $\mathbb{2} \otimes \mathbb{3}$ can be thought of a composite
system consisting of two independent unrelated subsystems. Each state of the
composite system therefore contains $\log (2 * 3) = \log 2 + \log 3$ bits which
is the sum of the information contained in each subsystem. If quantum field
theory is correct (as it so far seems to be) then information, during any
physical process, is neither created nor
destroyed. Landauer~\cite{Landauer:1961,bennett1985fundamental,Landauer},
Bennet~\cite{Bennett:1973:LRC,bennett2003notes,bennett2010notes},
Fredkin~\cite{fredkin1982conservative} and others made compelling arguments that
this physical principle induces a corresponding computational principle of
``conservation of information.'' In the context of finite types, generated from
the empty type $\zt$, the unit type $\ot$, and sums and products $\oplus$ and
$\otimes$, this principle states that the foundational (i.e., physical) notion
of computation is computation via type
isomorphisms~\cite{James:2012:IE:2103656.2103667} or type
equivalences~\cite{Carette2016}; these are both sound and complete with respect
to cardinality-preserving maps. The introduction, in $\pifrac$, of types
(groupoids) with fractional cardinalities introduces types with \emph{negative
  entropy}. For example, a type with cardinality $\frac{1}{8}$ has entropy
$\log{\frac{1}{8}} = -3$. It is natural to interpret this negative entropy just
like we interpret ``negative money,'' as a debt to be repaid by some other part
of the system. In fact, the zig-zag example can be viewed as modeling a credit
card transaction where the money is produced at the output site by generating a
corresponding debt that is reconciled at the input site.
